Having studied the vector space structure of $\R^n$ and the geometric behavior of linear maps, we will now turn our focus to studying  \textbf{non-linear}  functions from $\R^n$ to $\R^m$.  In particular, this will lead us to study the geometric and topological properties of $\R^n$. 

In section \ref{sec:multifunctions}, we will first explore functions $\R \to \R^n$ (\hyperref[vectorvaluedfunctions]{vector-valued functions}), as well as functions $\R^n \to \R$ (\hyperref[multivariablefunctions]{multivariable functions}).  These turn out to be enough to capture the behavior of \hyperref[generalmulti]{general multivariable functions} $\R^n \to \R^m$.

We will then study the notions of limits (section \ref{sec:limits}) and continuity  (section \ref{sec:continuity}) for these classes of functions, and see how it differs from limits and continuity in the single-variable case.


\section{Multivariable functions}\label{sec:multifunctions}

\subsection{Vector-valued functions}\label{vectorvaluedfunctions}

We will begin by studying vector-valued functions.  That is, maps of the form $f : \R \to \R^n$.  Note that the domain of $f$ is $\R$, so that the input of this function is a scalar.  On the other hand, the codomain of $f$ is $\R^n$, which means that the output is a point (position vector), hence the name vector-valued function. 

\begin{definition}
    A \textbf{vector-valued function}\define{vector-valued function} is a function $\bm{r}(t): \R \to \R^n$, given by
    $$\bm{r}(t) = \langle x_1(t),x_2(t), \cdots x_n(t) \rangle = x_1(t)\bm{e_1} + x_2(t)\bm{e_2} + \cdots + x_n(t)\bm{e_n}$$
    where all of the $x_i(t) : \R \to \R$ are single-variable functions.
    
    \vspace{1em}  
    $\bm{r}(t)$ is also called a \textbf{parametrization}\define{parametrization} (with \textbf{parameter} $t$).
    \end{definition}

In other words, vector-valued functions are precisely described by their component functions $x_i(t)$.  We will see that this will cause vector-valued functions $\bm{r}(t) : \R \to \R^n$ to behave extremely similarly to scalar functions $f : \R \to \R$.  In some sense, a vector-valued function $\bm{r}(t) : \R \to \R^n$ is simply a collection of $n$ single-variable scalar functions.


\begin{motivating}
How should we geometrically interpret vector-valued functions?
\end{motivating}

We should think of a vector-valued function $\bm{r}(t)$ as describing the motion of a particle in $\R^n$.  For example, let us think of the units of $t$ as begin seconds.  Then we can interpret the vector $\bm{r}(0)$ as describing the position of a particle at time $t=0$.  We can then think of the vector $\bm{r}(1)$ as describing the position of the particle 1 second after $t=0$; and similarly, the vector $\bm{r}(-1)$ describes the position of the particle 1 second \textbf{before} $t=0$.  

Let us turn to some examples:

\begin{example}
A scalar function $f : \R \to \R$ is a vector-valued function for $n=1$.
\end{example}


\begin{example}
Sketch the vector-valued function $\bm{c}(t) = \langle  -\sin(t), \cos(t) \rangle$ in $\R^2$.
\end{example}


\begin{example}\label{helix1}
Sketch the vector-valued function $\bm{h}(t) = \langle -\sin(t), \cos(t), t \rangle$ in $\R^3$.
\end{example}

\begin{example}\label{helix2}
Sketch the vector-valued function $\bm{r}(s) = \langle -\sin(5s), \cos(5s), 5s \rangle$ in $\R^3$.
\end{example}

\begin{motivating}
How do the two examples \hyperref[helix1]{$\bm{h}(t)$} and \hyperref[helix2]{$\bm{r}(s)$} differ from each other?
\end{motivating}

\begin{definition}

    Given a parametrization $\bm{r}(t) : \R \to \R^n$, we say that its \textbf{parametric curve}\define{parametric curve} is the set of points $$\{\bm{x} \in \R^n \ | \ \bm{x} = \bm{r}(t)\} \subset \R^n$$
    
\end{definition}

That is, we can geometrically interpret the parametric curve as being the path traced out by the parametrization $\bm{r}(t)$.  

\begin{example}
The parametric curve of $\bm{c}(t) = \langle  -\sin(t), \cos(t) \rangle$ is the circle of radius 1, centered at the origin.

That is, it is the set of points $\{(x,y) \ | \ x^2 + y^2=1 \}$
\end{example}


\begin{remark}
The two examples \hyperref[helix1]{$\bm{h}(t)$} and \hyperref[helix2]{$\bm{r}(s)$} show us that there are multiple possible parametrizations of the same curve.
\end{remark}

\begin{example}
The parametric curve of $\bm{h}(t) = \langle -\sin(t), \cos(t), t \rangle$ and $\bm{r}(s) = \langle -\sin(5s), \cos(5s), 5s \rangle$ is called a helix.

\begin{center}
    \begin{tikzpicture}
   \begin{axis}[view={60}{30},
                axis lines=center,axis on top,
                no marks,axis equal,
                xmin=-2,xmax=2,ymin=-2,ymax=2,zmin=-1,zmax=2,
                enlargelimits={upper=0.1}]
     \addplot3[
                UCLAblue,samples=100,
                domain=-1:4,
                y domain=0:0,
                ]
                ({cos(deg(2*pi*x))},{-1*sin(deg(2*pi*x))},{x});
  \end{axis}
\end{tikzpicture}    
    \end{center}

\end{example}


In other words, the parametrization describes how fast and in what direction you walk the path, whereas the parametric curve is the path itself.

\begin{motivating}
How can we sketch parametrizations and parametric curves?
\end{motivating}

Plotting points is something that a computer can do pretty easily; however, plotting points by hand is often a very inefficient way to study parametric curves.  We will discuss two tools that can help us more easily sketch and understand vector-valued functions in $\R^3$.

\subsubsection{Projections}

Given a vector-valued function $\bm{r}(t) = \langle x(t), y(t), z(t) \rangle$, if we isolate and study the behavior of any two components, then we can gain information about the parametrization.

\begin{definition}
Given a vector-valued function $\bm{r}(t) = \langle x(t), y(t), z(t) \rangle$, we can study the following \textbf{projections}:

\begin{itemize}
    \item The projection onto the $xy$-plane, which corresponds to $\bm{r}(t) = \langle x(t), y(t), 0 \rangle$
    \item The projection onto the $xz$-plane, which corresponds to $\bm{r}(t) = \langle x(t), 0, z(t) \rangle$
     \item The projection onto the $yz$-plane, which corresponds to $\bm{r}(t) = \langle 0, y(t), z(t) \rangle$
\end{itemize}

\end{definition}

That is, we can think of these as shadows of the vector-valued function $\bm{r}(t)$.

\begin{example}
Consider the helix parametrization $\bm{h}(t) = \langle -\sin(t), \cos(t), t \rangle$ in $\R^3$.

The projection onto the $xy$-plane corresponds to $\bm{r}(t) = \langle -\sin(t), \cos(t), 0 \rangle$.  This describes counterclockwise motion around the unit circle centered at the origin.  

The projection onto the $xz$-plane corresponds to $\bm{r}(t) = \langle -\sin(t), 0, t\rangle$.  This corresponds to a sine curve in the $xz$-plane.

The projection onto the $yz$-plane corresponds to $\bm{r}(t) = \langle 0, \cos(t), t \rangle$.  This corresponds to a sine curve in the $yz$-plane.

\fixthis{picture}

\end{example}





\subsubsection{Quadric surfaces in $\R^3$}

Another tool we can use to graph and sketch vector-valued functions in $\R^3$ are called quadric surfaces:  These are certain well-studied surfaces in $\R^3$, that have been completely classified.


\begin{definition}
A \textbf{quadric surface}\define{quadric surface} in $\R^3$ is defined as the set of points $$\{ (x, y, z) \ | \ Ax^2 + By^2 + Cz^2 + Dxy + Exz + Fyz + ax + by + cz + d = 0\}$$
\end{definition}

\begin{example}
A plane of the form $ax + by + cz + d = 0$ is a quadric surface.
\end{example}

Thus, studying these quadric surfaces is a generalization of our discussion of planes.  We are now allowing for non-linear behavior!

\fixthis{say more}

Let us investigate a few non-linear examples:

\begin{example}
Let us consider the quadric surface $C$ given by  $$C := \{ (x, y, z) \ | \ x^2 + y^2 = 1\}$$

Observe that if we restrict our attention to the points that lie in both $C$ and in the plane $z = 0$, we see that these points can be described precisely as a unit circle centered at the origin, in the $z=0$ plane.

$$\{ (x, y, 0) \ | \ x^2 + y^2 = 1\}$$

Furthermore, this same picture holds if we instead consider any plane of the form $z=k$!  We again see a unit circle centered at the origin, but now in the $z=k$ plane.

$$\{ (x, y, k) \ | \ x^2 + y^2 = 1\}$$

Thus, the surface $C$ consists of a collection (one for each $k \in \R$) of unit circles centered at the origin, stacked on top of each other.  In other words, $C$ is an infinite cylinder of radius 1.
\end{example}

\fixthis{define cylinders}

We can use this technique in general to study the solutions to an arbitrary equation in $\R^3$ \fixthis{ref later}

\begin{definition}
    The \textbf{trace in the plane $P$}\define{trace} of a subset $S \subset \R^3$ is the intersection of $S$ with $P$.  That is, 
    $$S \cap P = \{\bm{x} \in \R^3 \ | \ \bm{x} \in S \textnormal{and} \bm{x} \in P\}$$
    
    We typically study $x$-traces, $y$-traces, and $z$-traces - that is, traces in the planes $x=k$, $y=k$, and $z=k$ respectively.
    
\end{definition}

We can think of traces as being slices of the graph. \fixthis{say more}


\begin{theorem}\label{quadraticcylinders}
Up to rotation and translation, there are 3 types of quadratic cylinders:

\begin{itemize}
    \item Parabolic cylinders, which have standard form $$y = \left(\frac{x}{a}\right)^2$$
    \item Elliptic cylinders, which have standard form $$\left(\frac{x}{a}\right)^2 + \left(\frac{y}{b}\right)^2 = 1$$
    \item Hyperbolic cylinders , which have standard form $$\left(\frac{x}{a}\right)^2 - \left(\frac{y}{b}\right)^2 = 1$$
\end{itemize}

\end{theorem}







\begin{theorem}
Up to rotation and translation, there are 8 types of quadric surfaces:

\begin{itemize}
    \item Planes, which have standard form $$ax + by + cz = d$$
    \item Quadratic cylinders (described in theorem \ref{quadraticcylinders}).
    \item Elliptic paraboloids, which have standard form $$z = \left(\frac{x}{a}\right)^2 + \left(\frac{y}{b}\right)^2$$
    \item Hyperbolic paraboloids, which have standard form $$z = \left(\frac{x}{a}\right)^2 - \left(\frac{y}{b}\right)^2$$
    \item Ellipsoids, which have standard form $$\left(\frac{x}{a}\right)^2 + \left(\frac{y}{b}\right)^2  + \left(\frac{z}{c}\right)^2 = 1$$
    \item Elliptic cones, which have standard form $$\left(\frac{z}{c}\right)^2  = \left(\frac{x}{a}\right)^2 + \left(\frac{y}{b}\right)^2$$
    \item Hyperboloids of one sheet, which have standard form $$\left(\frac{z}{c}\right)^2 +1 = \left(\frac{x}{a}\right)^2 + \left(\frac{y}{b}\right)^2  = \left(\frac{z}{c}\right)^2 + 1 $$
    \item Hyperboloids of two sheets, which have standard form $$\left(\frac{z}{c}\right)^2 - 1 = \left(\frac{x}{a}\right)^2 + \left(\frac{y}{b}\right)^2$$
\end{itemize}
\end{theorem}

That is, given any quadric surface, corresponding to a quadratic equation $Ax^2 + By^2 + Cz^2 + Dxy + Exz + Fyz + ax + by + cz + d = 0$, one can rotate it and translate it so that the quadratic equation can be written in the standard form above.



\begin{example}
The unit sphere of radius 1 can be described as the set of vectors $\{\bm{v} \in \R^3 \ | \ ||\bm{v}|| = 1\}$.  Equivalently, the unit sphere is the set of points $$\{ (x, y, z) \ | x^2 + y^2 + z^2 =1\}$$

Thus, the unit sphere of radius 1 is an example of an ellipsoid!
\end{example}

\begin{example}
Sketch and describe the surface $$x = \left(\frac{y-1}{4}\right)^2 + \left(\frac{z}{3}\right)^2 $$
\end{example}

\begin{example}
Describe and sketch the the quadric surface $S$ given by $$\frac{9(x-1)^2}{4}+4z^2 =9y^2 + 36$$
\end{example}


\begin{motivating}
How can we use this to sketch vector-valued functions?
\end{motivating}


Given a vector-valued function $\bm{r}(t) = \langle x(t), y(t), z(t) \rangle$, if we can find a relationship between the components of the coordinate vectors, then we know that $\bm{r}(t)$ must lie on the quadric surface.

\begin{example}
Consider the helix parametrization $\bm{h}(t) = \langle -\sin(t), \cos(t), t \rangle$ in $\R^3$.

Observe that the components of $\bm{h}(t)$ satisfy the equation $x(t)^2 + y(t)^2 = 1$.  This tells us indeed that the helix lies on the cylinder $x^2 + y^2 = 1$.

\fixthis{draw picture}
\end{example}



\subsection{Multivariable functions}\label{multivariablefunctions}

We now to turn to studying functions of the form $f : \R^n \to \R$:

\begin{definition}
A \textbf{multivariable function}\define{multivariable function} is a function $f : \R^n \to \R$.  We sometimes say that $f$ is a \textbf{function of $n$ variables}\define{function of $n$ variables}.
\end{definition}

\begin{example}
A scalar function $f : \R \to \R$ is a function of 1 variable.
\end{example}

\begin{example}
    The distance between the origin and a point $P = (x_1, \cdots, x_n)$  is a multivariable function $d : \R^n \to \R$ defined by 
    $$d(x_1, \cdots, x_n) = \sqrt{\sum_{i=1}^n x_i^2}$$
\end{example}

\begin{definition}
    Consider a subset $D \subset \R^n$.  The \textbf{indicator function} $1_D$ is the piecewise function defined by
    $$1_D(\bm{x}) := \left\{
		\begin{array}{ll}
			1 & \text{ if } \bm{x} \in D \\
			0 & \text{ if } \bm{x} \notin D
		\end{array}
		\right.$$
    
    \end{definition}



Just like with single-variable functions, we might be interested in functions that are defined only on a subset $D \subseteq \R^n$.  Nevertheless, the following construction shows us that it is sufficient to study the theory of multivariable functions $f : \R^n \to \R$.

\begin{example}
Let $D \subset \R^n$.  Then a function $f : D \to \R$ can be extended to a function of $n$ variables $\Tilde{f} : \R^n \to \R$ in the following way:
$$\Tilde{f}(\bm{x}) := \left\{
		\begin{array}{ll}
			f(\bm{x}) & \text{ if } \bm{x} \in D \\
			0 & \text{ if } \bm{x} \notin D
		\end{array}
		\right.$$

That is, to every point $\bm{x}=(x_1,\cdots,x_n)$ in the domain $D$, we assign the real number $f(\bm{x}) \in \R$.  To every other point $\bm{x} \notin D$, we assign the value $0$.  In other words,  $\Tilde{f}(\bm{x}) = f(\bm{x})1_D(\bm{x})$.  

\end{example}

\begin{motivating}
How should we geometrically interpret multivariable functions?
\end{motivating}

Recall that given a single variable function $f: \R \to \R$, we can draw its graph as the set of points $$\{(x, f(x)) \} \subseteq \R^2$$  We can generalize this to graph multivariable functions $f : \R^n \to \R$:

\begin{definition}
Given a function $f: \R^n \to \R$, its \textbf{graph}\define{graph of a multivariable function} is the following subset of $\R^{n+1}$:
$$\Gamma_f := \{(x_1,\cdots, x_n,f(x_1, \cdots, x_n)) \} \subset \R^{n+1}$$ 

In other words, the graph is given by the equation $$x_{n+1} = f(x_1, \cdots, x_n)$$ in $\R^{n+1}$.
\end{definition}

We can easily sketch and visualize graphs of functions with $2$ variables:

\begin{definition}
    The \textbf{graph} of a function of two variables $f(x,y)$ is the surface in $\R^3$ consisting of the set of points $(x,y,z)$ that are solutions to the equation
    $$z = f(x,y)$$
\end{definition}

\begin{motivating}
How can we sketch the graph of a function of two variables?
\end{motivating}

To sketch graphs in $\R^3$, we can use traces!

\begin{definition}
    The \textbf{trace in the plane $P$}\define{trace} of a graph $\Gamma \subset \R^3$ is the intersection of $\Gamma$ with $P$.  
    
    We typically study $x$-traces, $y$-traces, and $z$-traces - that is, traces in the planes $x=k$, $y=k$, and $z=k$ respectively.
    
\end{definition}

We can think of traces as being slices of the graph. \fixthis{say more}

\begin{definition}
    The \textbf{level curves}\define{level curves} (isoclines, contour map) of a function of two variables $f(x,y)$ are the $z$-traces of the graph $z = f(x,y)$.
\end{definition}


\fixthis{picture}



\begin{example}
Sketch the graph of $f(x,y) = x^2 + y^2$.
\end{example}

\begin{example}
Sketch the graph of $f(x,y) = \sqrt{9-x^2 - y^2}$.
\end{example}



\subsubsection{Implicit functions}

From the previous section, we observed that some surfaces, such as the sphere in $\R^3$, are not graphs of multivariable functions.  That is, the sphere cannot be described as $z = f(x,y)$ for $f(x,y)$ a function of 2 variables. (As we will see later, they are graphs of multivariable functions \textbf{locally}).

Nevertheless, we can still describe them using multivariable functions implicitly!

\begin{definition}
Given a multivariable function $G(x_1, \cdots, x_n) : \R^n \to \R$, its \textbf{vanishing locus}\define{vanishing locus} is the set of points 
$$\{(x_1, \cdots, x_n) \ | \ G(x_1, \cdots, x_n) = 0\}$$
\end{definition}


\begin{example}
The unit sphere in $\R^3$ is the vanishing locus of the function $$G(x,y,z) = x^2 + y^2 +z^2-1$$
\end{example}

\begin{example}
All quadric surfaces are the vanishing loci of the general quadratic equation $$Q(x,y,z) = Ax^2 + By^2 + Cz^2 + Dxy + Exz + Fyz + ax + by + cz + d$$
\end{example}

We can also describe graphs of functions as vanishing loci!

\begin{example}
The graph of a multivariable function $x_{n+1}=f(x_1, \cdots, x_n)$ is the vanishing locus of the function $$G(x_1, \cdots, x_{n+1}) = f(x_1, \cdots, x_n) -x_{n+1}$$
\end{example}

Note that the vanishing locus of $G : \R^n \to \R$ is a subset of $\R^n$, whereas the graph of a function $f : \R^n \to \R$ is a subset of $\R^{n+1}$!  Again, this is analogous to the difference between the kernel of a linear map, versus the image of a linear map.






\subsection{General Multivariable functions}\label{generalmulti}

We now to turn to the general case:

\begin{definition}
A \textbf{general multivariable function}\define{general multivariable function} is a function $f : \R^n \to \R^m$, defined by 
$$f(x_1, \cdots, x_n) = \big\langle f_1(x_1, \cdots, x_n), f_2(x_1, \cdots, x_n), \cdots, f_m(x_1, \cdots, x_n) \big\rangle$$

\end{definition}

That is, a general multivariable function is a vector-valued function, whose components are multivariable functions!

\begin{example}
A vector-valued function is a general multivariable function $f : \R \to \R^m$.
\end{example}

\begin{example}
A multivariable function is a general multivariable function $f : \R^n \to \R$.
\end{example}


\subsection{Exercises}

\begin{problem}{Cycloid}

Consider the vector-valued function $\bm{r}(t) = \langle t-\sin(t), 1-\cos(t) \rangle$, which parametrizes a curve called a \textbf{cycloid}.  Sketch $\bm{r}(t)$ from $0$ to $2\pi$.

\end{problem}

\begin{problem}{swirly}
Sketch and describe the vector-valued function $\bm{r}(t) = \langle e^{-t}\sin(t), e^{-t}\cos(t), e^{-t} \rangle$.  

\begin{subproblems}
\item How many times does $\bm{r}(t)$ intersect a sphere of radius 1?
\end{subproblems}

\end{problem}

\begin{problem}{conicalhelix}
Sketch and describe the vector-valued function $\bm{r}(t) = \langle t\sin(t), t\cos(t), t \rangle$.  
\end{problem}

\begin{problem}{cylinderslice}
Find a parametrization of the intersection of the cylinder $x^2 + y^2 = 1$ and the plane $x + y + z = 1$.

Sketch and describe the parametric curve.
\end{problem}

\begin{problem}{cylinderparbola}
Find a parametrization of the intersection of the cylinder $x^2 + y^2 = 1$ and the parabolic cylinder $z = 4x^2$

Sketch and describe the parametric curve.
\end{problem}

\begin{problem}{Vistani1}
Vistani's curve is the intersection of the surfaces $x^2 + y^2 = z^2$ and $y=z^2$.  Find a parametrization of Vistani's curve.
\end{problem}

\begin{problem}{Vistani2}
Show that Vistani's curve lies on the unit sphere of radius 1, centered at $(0,1,0)$.
\end{problem}

\begin{problem}{inaplane}
    Show that the vector-valued function $\bm{r}(t) : \langle t^2-1, t-2t^2, 4-6t \rangle$ lies in a plane.  What is the equation of the plane?
\end{problem}

We say that two lines $\bm{r_1}(t)$ and $\bm{r_2}(s)$ \textbf{intersect} if there is a point $P$ lying on both curves.  We say that $\bm{r_1}(t)$ and $\bm{r_2}(s)$ \textbf{collide} if there exists some $t_0$ such that $\bm{r_1}(t_0) = \bm{r_2}(t_0)$.

\begin{problem}{colvsint}
What is the difference between a point of intersection and a point of collision?
\end{problem}

\begin{problem}{collision}
Consider the vector-valued functions $$\bm{r_1}(t) = \langle t^2 - 3, -1, t^2 \rangle \qquad \bm{r_2}(s) = \langle 2s, 2s+1, 2s+3 \rangle$$

\begin{subproblems}
\item Do $\bm{r_1}(t)$ and $\bm{r_2}(s)$ \textbf{collide}?  If so, find the point(s) of collision.
\item Find all values of $t$ and $s$ where the curves of $\bm{r_1}(t)$ and $\bm{r_2}(s)$ \textbf{intersect}.
\end{subproblems}

\end{problem}










\begin{problem}{traces1}
Sketch traces of the surface $x^2 + (\frac{y}{4})^2 + z^2 = 1$.
\end{problem}

\begin{problem}{quadricsurface1}
Classify and sketch the quadric surface $S$ given by $$\left(\frac{x}{2}\right)^2+\left(\frac{y}{5}\right)^2 =5z^2 - 1$$
\end{problem}

\begin{problem}{quadricsurface2}
Classify and sketch the quadric surface $S$ given by  $$9y^2+4\left(z-1\right)^2 =\frac{9x^2}{4} + 36$$
\end{problem}

\begin{problem}{traces2}
Sketch traces of the function $f(x,y) = \frac{y}{x^2}$.
\end{problem}

\begin{problem}{traces3}
Sketch traces of the function $f(x,y) = \frac{1}{x^2 + y^2 + 1}$.
\end{problem}

\begin{problem}{traces3}
Sketch traces of the function $f(x,y) = \cos(x-y)$.
\end{problem}

\begin{problem}{tracesinverse}
Consider the following equations:
    \begin{center}
    \def\arraystretch{1.5}
        \begin{tabular}{lc}
(i) & $\left(\frac{y}{2}\right)^2 =\left(\frac{x}{4}\right)^2 + \frac{8}{9}$                                         \\
(ii) & $\left(\frac{y}{2}\right)^2 =\left(\frac{x}{4}\right)^2 + 1$     \\
(iii) & $\left(\frac{y}{2}\right)^2+\left(\frac{z-1}{3}\right)^2 = 1$ 
\end{tabular}
    \end{center}
    
    \begin{subproblems}
    \item If a quadric surface $S$ has $(i)$ and $(ii)$ as traces, then can we determine $S$?
    \item If a quadric surface $S$ has $(i)$, $(ii)$, and $(iii)$ as traces, then can we determine $S$?
    \item If $(i)$ is a trace of a quadric surface $9y^2+4\left(z-1\right)^2 =\frac{9x^2}{4} + 36$, what plane could the trace be in?
    \item If $(iii)$ is a trace of a quadric surface $9y^2 + 4(z-1)^2 + 9(x-1)^2 = 36$, what plane could the trace be in?
    \end{subproblems}
    
\end{problem}


\begin{problem}{levelcurves1}
Sketch the level curves of the multivariable function $f(x,y) = |x| + |y|$
\end{problem}




\section{Limits of sequences}\label{sec:limits}

Now that we have a good understanding of multivariable functions, let us turn to rigorously studying limits and continuity in multivariable calculus.  We will first study limits of sequences (\hyperref[limseq]{of vectors}), and then turn to limits of functions (both \hyperref[limvectorval]{vector-valued} and \hyperref[limmulti]{multivariable} functions).

First, recall the notion of a limit of sequences of real numbers:

\begin{definition}
Let $\{a_k\}$ be a sequence of real numbers.  We say that the sequence \textbf{converges} if the following holds:

For all $\varepsilon>0$, there exists $M$ such that for all $m > M$, $$|a_m - L | < \varepsilon$$
    
We say $L$ is the \textbf{limit} of the sequence $\{a_k\}$.  If no such $L$ exists, we say $\{a_k\}$ \textbf{diverges}\define{sequence divergence}.
\end{definition}

In other words, the limit of a sequence $\{a_k\}$ is $L$ if for every small value $\varepsilon$, we can find an $M$ such that $a_m$ is $\varepsilon$-close to $L$ for $m > M$.



\begin{example}
Consider the sequence $\{\frac{1}{k^3 + 1}\}$.  Then $\lim_{k \to \infty} \frac{1}{k^3 + 1} = 0$.

\begin{proof}
Let $\varepsilon > 0$.  We wish to find $M$ such that for $m > M$, $|a_m - 0|  = a_m < \varepsilon$.

Let us choose $M > (\frac{1}{\varepsilon}-1)^{\frac{1}{3}}$.  Then if $m > M$, we have that $$m^3 > M^3 >  (\frac{1}{\varepsilon}-1)$$
Thus, $$m^3 +1  > \frac{1}{\varepsilon}$$
We can then deduce that for $m > M$,  $$\varepsilon > \frac{1}{m^3 +1} =a_m$$
That is, we have shown that $\lim_{k \to \infty} \frac{1}{k^3 + 1} = 0$.


\end{proof}

\end{example}





\begin{motivating}
Why do we need to use $\varepsilon$ when talking about limits?
\end{motivating}

Ideally, what we would like to say is that the sequence $\{a_k\}$ will eventually reach its limit, $L$.  However, if we consider the sequence $$\left\{\frac{1}{k^3 + 1}\right\}$$
We can observe that $\lim_{k\to \infty}\frac{1}{k^3 + 1} = 0$, but $\frac{1}{k^3 + 1} > 0$.

Thus, since the sequence $\{a_k\}$ might not ever attain its limit $L$, we need a different way to describe the limit:

\begin{proposition}\label{epsilonenough}
    Let $x \in \R_{\geq0}$.  Then the following are equivalent:
    
    \begin{enumerate}
        \item For all $\varepsilon >0$,  $x < \varepsilon$
        \item $x = 0$.
    \end{enumerate}
    \end{proposition}

As a consequence, to show that $a = b$, it is equivalent to show that there is no number between $a$ and $b$!


\begin{example}
Consider the sequence $\{\sum_{i=1}^k \frac{9}{10^i}\} = \{0.9, 0.99, 0.999, \cdots \}$.  Then $\lim_{k \to \infty} a_k = 1$.

In other words, the notion of limits explains why $0.\overline{999} = 1$.
\end{example}

Observe that in our definition, we use the term "\textbf{the} limit", which should imply that the limit of a sequence of real numbers is unique.  We verify that this is true:



\begin{proposition}[Limits are unique]\label{limsequnique1}
    Suppose $\{a_k\}$ is a sequence of real numbers that converges to both $a$ and $b$.  Then $a = b$.
    \end{proposition}

\begin{proof}
By proposition \ref{epsilonenough}, it is equivalent to show that for all $\varepsilon > 0$, $|a-b| < \varepsilon$.  Thus, let $\varepsilon > 0$.  

Since $\lim_{k \to \infty} a_k = a$, we know that there exists $M_a$ such that for all $m > M_a$, $|a_m -a| < \frac{\varepsilon}{2}$. Similarly, since $\lim_{k\to \infty} a_k = b$, we know that there exists $M_b$ such that for all $m > M_b$, $|a_m -b| < \frac{\varepsilon}{2}$.  Set $M = \textnormal{max}\{M_A, M_B\}$

Observe that $|a-b| = |a - a_m + a_m - b|$, and we can apply the triangle inequality to observe that for all $m > M$,
\begin{align*}
    |a-b| &= |a - a_m + a_m - b| \\
    & \leq |a-a_m| + |b-a_m| \\
    &< \frac{\varepsilon}{2} + \frac{\varepsilon}{2} = \varepsilon
\end{align*}

Thus, we have proven that for all $\varepsilon > 0$, $|a-b| < \varepsilon$.  Therefore, $a = b$, as desired.

\end{proof}


We can now rigorously prove the limit properties that we've seen in single variable calculus.


\begin{theorem}[Properties of Limits in $\R$]
    Let $\{a_k\}$ and $\{b_k\}$ be sequences in $\R$ that converge to $a$ and $b$, respectively. Then
    
    \begin{enumerate}
        \item $\lim_{k \to \infty}(a_k + b_k) = a + b$
        \item $\lim_{k \to \infty}\left(ca_k\right) = ca$ for all $c \in \R$.
        \item $\lim_{k \to \infty}\left(a_kb_k\right) = ab$.
        \item If $\{b_k\}, b \neq 0$, then $\lim_{k \to \infty}\frac{1}{b_k} = \frac{1}{b}$.
        \item If $\{b_k\}, b \neq 0$, then $\lim_{k \to \infty}\frac{a_k}{b_k} = \frac{a}{b}$.
    \end{enumerate}
    
    
    \end{theorem}
    
    
    
    
    
    

\begin{theorem}[Squeeze theorem for sequences]
    Let $\{a_k\}$, $\{b_k\}$, and $\{c_k\}$ be sequences in $\R$ such that there exists an $M$ such that $a_m \leq b_m \leq c_m$ for all $m > M$.  
    
    Then if $\lim_{k \to \infty}a_k = \lim_{k \to \infty}c_k = L$, then $\lim_{k \to \infty}b_k = L$.
    
\end{theorem}    
    
\begin{corollary}
    If $\lim_{k \to \infty} |a_k| = 0$, then $\lim_{k \to \infty}a_k = 0$.
\end{corollary}
    
    
    
    
    



\subsection{Limits of sequences of vectors}\label{limseq}

\begin{motivating}
How do we generalize the notion of sequences in $\R$ to sequences of vectors in $\R^k$?
\end{motivating}


Recall that the idea of a limit of a sequence of real numbers was the following:  $L$ is the limit of a sequence $\{a_n\}$ if for every small value $\varepsilon$, we can find an $N$ such that $a_n$ is $\varepsilon$-close to $L$ for $n > N$. 

The same idea should hold true for vectors in $\R^n$ - we We say that two points $\bm{P}, \bm{Q} \in \R^n$ are close to each other if the distance $||\bm{P-Q}||$ between them is small!


    \begin{definition}
    Let $\{\bm{a_n}\}$ be a sequence of vectors in $\R^k$.  We say that the sequence $\{\bm{a_n}\}$ \textbf{converges} to the vector $\bm{L} \in \R^k$ if:
    
    \vspace{1em}
    For all $\varepsilon>0$, there exists $M$ such that for all $m > M$, $$||\bm{a_m} - \bm{L} || < \varepsilon$$
    
    We say $\bm{L}$ is the limit of the sequence $\{\bm{a_n}\}$. If no such $\bm{L}$ exists, we say that $\{\bm{a_n}\}$ diverges.
\end{definition}

Another way to state the definition of limits is through topology:

\begin{definition}
    Let $\bm{P} \in \R^n$.  \textbf{The open ball of radius $\varepsilon$ around $\bm{P}$}\define{open ball}, denoted $B_{\varepsilon}(\bm{P})$, is the set of points 
    
    $$B_{\varepsilon}(\bm{P}) := \{ \bm{x} \in \R^n \ | \ ||\bm{x} - \bm{P} || < \varepsilon\}$$
    
    \end{definition}

    That is, $\lim_{n\to\infty} \bm{a_n} = \bm{L}$ if for every $\varepsilon > 0$, there is some $M$ such that $\bm{a_m} \in B_{\varepsilon}(\bm{L})$.

    \fixthis{picture}


Once again, we should confirm that the terminology "\textbf{the} limit" is correct.  The key idea of the proof is the same as in proposition \ref{limsequnique1}.

\begin{proposition}[Limits are unique]
    Suppose $\{\bm{a_n}\}$ converges to both $\bm{a}$ and $\bm{b}$ in $\R^n$.  Then $\bm{a} = \bm{b}$.
    \end{proposition}

\begin{proof}
By proposition \ref{epsilonenough}, it is equivalent to show that for all $\varepsilon > 0$, $||\bm{a}-\bm{b}|| < \varepsilon$.  Thus, let $\varepsilon > 0$.  

Since $\lim_{k \to \infty} \bm{a_k} = \bm{a}$, we know that there exists $M_a$ such that for all $m > M_a$, $||\bm{a_m} -\bm{a}|| < \frac{\varepsilon}{2}$. Similarly, since $\lim_{k\to \infty} \bm{a_k} = \bm{b}$, we know that there exists $M_b$ such that for all $m > M_b$, $||\bm{a_m} -\bm{b}|| < \frac{\varepsilon}{2}$.  Set $M = \textnormal{max}\{M_A, M_B\}$

Observe that $||\bm{a}-\bm{b}|| = ||\bm{a}-\bm{a_m}+\bm{a_m}-\bm{b}||$, and we can apply the triangle inequality to observe that for all $m > M$,
\begin{align*}
    ||\bm{a}-\bm{b}|| &= ||\bm{a}-\bm{a_m}+\bm{a_m}-\bm{b}|| \\
    & \leq ||\bm{a}-\bm{a_m}||+||\bm{a_m}-\bm{b}|| \\
    &< \frac{\varepsilon}{2} + \frac{\varepsilon}{2} = \varepsilon
\end{align*}

Thus, we have proven that for all $\varepsilon > 0$, $||\bm{a}-\bm{b}|| < \varepsilon$.  Therefore, $\bm{a} = \bm{b}$, as desired.

\end{proof}





\begin{theorem}[Limits in $\R^m$ are determined componentwise]\label{limseqcomponentwise}
    Let $\{\bm{a_n}\}$ be a sequence in $\R^m$, where $\bm{a_n} = \langle a_{n,1},\  a_{n,2},\  \cdots \ a_{n,m} \rangle$.
    
    \vspace{1em}
    
    Then $\{\bm{a_n}\}$ converges to $\bm{a} = \langle a_1, \cdots, a_m \rangle$ if and only if $\{a_{n,i}\}$ converges to $a_i$ for all $1 \leq i \leq m$.
    
    
    \end{theorem}

\begin{proof}
The key idea is the following inequalities $|a_{n,i} - a_i| \leq ||\bm{a_{n,i}} - \bm{a}|| \leq \sum_{i=1}^m |a_{n,i} - a_i|$.
\end{proof}



    \begin{corollary}
    Let $\{\bm{a_n}\}$ and $\{\bm{b_n}\}$ be sequences in $\R$ that converge to $\bm{a}$ and $\bm{b}$, respectively. Then
    
    \begin{enumerate}
        \item $\lim_{n \to \infty}(\bm{a_n} + \bm{b_n}) = \bm{a + b}$
        \item $\lim_{n \to \infty}\left(c\bm{a_n}\right) = c\bm{a}$ for all $c \in \R$.
    \end{enumerate}
    
    
    \end{corollary}







\subsection{Divergent sequences}




\begin{motivating}
Do limits of sequences always exist?
\end{motivating}

There are two main types of sequences that diverge:

    \begin{example}
    The sequence $\{(-1)^n\}$ in $\R$ diverges.
    \end{example}
    
    \begin{example}
    The sequence $\{n\}$ in $\R$ diverges.
    \end{example}
    
    We will investigate the behavior of both of the sequences.


\begin{definition}
    Let $\{\bm{a_n}\}$ be a sequence of vectors in $\R^k$.  A \textbf{subsequence} of $\{\bm{a_n}\}$ is a sequence $\{\bm{b_i}\}$, where $$\bm{b_i} = \bm{a_{n_i}}$$
    such that $n_1 < n_2 < \cdots < n_i < \cdots$.
    \end{definition}

    
    \begin{theorem}
    Let $\{\bm{a_n}\}$ be a sequence of vectors in $\R^k$ such that $\{\bm{a_n}\}$ has two different subsequences $\{\bm{a_{n_i}}\}$ and $\{\bm{a_{n_j}}\}$ that converge to two different limits.
    
    Then  $\{\bm{a_n}\}$ diverges.
    \end{theorem}
    
    
     \begin{example}
    The sequence $\{(-1)^n\}$ has the subsequences $$\{(-1)^{2n}\} \textnormal{ and } \{(-1)^{2n+1}\}$$ which converge to $1$ and $-1$, hence $\{(-1)^n\}$ diverges.
    \end{example}

\begin{theorem}[Convergent sequences are bounded]

Let $\{\bm{a_n}\}$ be a sequence in $\R^k$ such that $\lim_{n \to \infty}\bm{a_n} = \bm{L}$.

\vspace{1em}

Then the sequence $\{\bm{a_n}\}$ is \textbf{bounded}.  That is, there exists a number $A \in \R$ such that $||\bm{a_n}|| \leq A$ for all $n$.
\end{theorem}
    
    \begin{proof}
    \fixthis{picture}
    \end{proof}
    
    
    \begin{corollary}
    If $\{\bm{a_n}\}$ be a sequence in $\R^k$ that is unbounded, then $\{\bm{a_n}\}$ diverges.
    \end{corollary}

\begin{example}
    The sequence $\{n\}$ in $\R$ is unbounded, hence it diverges.
    \end{example}




\begin{theorem}\label{divergencecriteria}
Let $\{\bm{a_n}\}$ be a sequence of vectors in $\R^k$.  If $\{\bm{a_n}\}$ has a subsequence $\{\bm{a_{n_i}}\}$ that diverges, then $\{\bm{a_n}\}$ diverges.

\end{theorem}







\section{Limits of functions}




Now let us turn to limits of functions!  Let us first recall the notion of limits of single-variable functions:


\begin{definition}
    A function $f :\R \to \R$ has the \textbf{limit} $b$ at $a$ if
    
    \vspace{1em}
    For all $\varepsilon>0$, there exists $\delta > 0$ such that for all $\bm{x} \in \R$, $$0 < |x-a| < \delta \qquad \textnormal{implies} \qquad |f(x) - b | < \varepsilon$$
    
    We write $\lim_{x \to a}f(x) = b$.
    
    \end{definition}


Since $\R$ is a well-ordered field (in other words, since ), we can turn this into a statement about left limits and right limits.

\fixthis{left and right picture}









\begin{definition}[Limit of a general multivariable function]
    A function $f : \R^n \to \R^m$ has the \textbf{limit} $\bm{b}$ at $\bm{a}$ if
    
    \vspace{1em}
    For all $\varepsilon>0$, there exists $\delta > 0$ such that for all $\bm{x} \in \R^n$, $$0 < ||\bm{x} - \bm{a} || < \delta \qquad \textnormal{implies} \qquad ||f(\bm{x}) - \bm{b} || < \varepsilon$$
    
    We write $\lim_{\bm{x} \to \bm{a}}f(\bm{x}) = \bm{b}$.
    
    \end{definition}


    \begin{proposition}[Limits of functions are unique]
    Let $f: \R^n \to \R^m$, and suppose $\lim_{\bm{x} \to \bm{a}}f(x) = \bm{b_1}$ and $\lim_{\bm{x} \to \bm{a}}f(x) = \bm{b_2}$.  Then $\bm{b_1} = \bm{b_2}$.
    \end{proposition}

We will study this for the two different classes of functions (\hyperref[limvectorval]{vector-valued functions} $\R \to \R^n$ and \hyperref[limmulti]{multivariable functions} $\R^n \to \R$)  we've seen.














\subsection{Limits of vector valued functions}\label{limvectorval}

\begin{definition}[Limit of a vector-valued function]
    A function $f : \R \to \R^m$ has the \textbf{limit} $\bm{b}$ at $a$ if
    
    \vspace{1em}
    For all $\varepsilon>0$, there exists $\delta > 0$ such that for all $x \in \R$, $$0 < |x-a| < \delta \qquad \textnormal{implies} \qquad ||f(\bm{x}) - \bm{b} || < \varepsilon$$
    
    \end{definition}


\begin{theorem}[Limits of vector-valued functions are determined componentwise]
    For any vector valued function $\bm{r}(t) = \langle x_1(t), \cdots, x_n(t) \rangle$, $$\lim_{t \to a} \bm{r}(t) = \langle x_1, \cdots, x_n \rangle$$ if and only if for all $1 \leq i \leq n$, $$\lim_{t \to a} x_i(t) = x_i$$
    
    
    \end{theorem}

    \begin{proof}
    The proof is essentially the same as the proof of theorem \ref{limseqcomponentwise}:
    
    The key idea is again the following inequalities $|x_i(t) - x_i| \leq ||\bm{r}(t) - \langle x_1, \cdots, x_n \rangle|| \leq \sum_{i=1}^m |x_i(t) - x_i|$.
    
    
    
    \end{proof}

    
\fixthis{picture}


\begin{theorem}
    Let $\bm{r},\bm{s} : \R \to \R^m$ be vector-valued functions. Suppose that $\lim_{t \to a}\bm{r}(t)$ and $\lim_{t \to a}\bm{s}(t)$ exist. Then 
    
    \begin{enumerate}
        \item \textbf{Sum Law:} $$\lim_{t \to a} \left(\bm{r}(t) + \bm{s}(t))\right) = \lim_{t \to a} \bm{r}(t) + \lim_{t \to a} \bm{s}(t)$$
        \item \textbf{Scalar Multiple Law:} $$\lim_{t \to a} \lambda(\bm{r}(t)) = \lambda \left(\lim_{t \to a} \bm{r}(t) \right)$$
    \end{enumerate}
    
    
    \end{theorem}



\subsection{Limits of multivariable functions}\label{limmulti}

\begin{definition}[Limit of a multivariable function]
    A function $f : \R^n \to \R$ has the \textbf{limit} $b$ at $\bm{a}$ if
    
    \vspace{1em}
    For all $\varepsilon>0$, there exists $\delta > 0$ such that for all $\bm{x} \in \R^n$, $$0 < ||\bm{x} - \bm{a} || < \delta \qquad \textnormal{implies} \qquad |f(\bm{x}) - b | < \varepsilon$$
    \end{definition}

\begin{theorem}\label{limmultirules}
     Let $f,g : \R^n \to \R$ be functions of $n$ variables. Suppose that $\lim_{\bm{x} \to \bm{P}}f(\bm{x})$ and $\lim_{\bm{x} \to \bm{P}}g(\bm{x})$ exist. Then 
    
    \begin{enumerate}
        \item \textbf{Sum Law:} $$\lim_{\bm{x} \to \bm{P}} \left(f(\bm{x}) + g(\bm{x})\right) = \lim_{\bm{x} \to \bm{P}} f(\bm{x}) + \lim_{\bm{x} \to \bm{P}} g(\bm{x})$$
        \item \textbf{Scalar Multiple Law:} $$\lim_{\bm{x} \to \bm{P}} \lambda(f(\bm{x})) = \lambda \left(\lim_{\bm{x} \to \bm{P}} f(\bm{x}) \right)$$
        \item \textbf{Product Law:} $$\lim_{\bm{x} \to \bm{P}} \left(f(\bm{x})g(\bm{x})\right) = \left(\lim_{\bm{x} \to \bm{P}}f(\bm{x})\right)\left(\lim_{\bm{x} \to \bm{P}} g(\bm{x})\right)$$
        \item \textbf{Quotient Law:} If $\lim_{\bm{x} \to \bm{P}} g(\bm{x})\neq 0$, $$\lim_{\bm{x} \to \bm{P}} \frac{f(\bm{x})}{g(\bm{x})} = \frac{\lim_{\bm{x} \to \bm{P}}f(\bm{x})}{\lim_{\bm{x} \to \bm{P}} g(\bm{x})}$$
    \end{enumerate}
    
    
    \end{theorem}

This theorem allows us to reduce to studying limits of single-variable functions.
    
    \begin{example}
Let $f(x,y) = x$.  Find $$\lim_{(x,y) \to (a,b)}f(x,y)$$
\end{example}    


\begin{example}
Let $f(x,y) = \frac{x\sin(y)}{e^x+1}$.  Find $$\lim_{(x,y) \to (a,b)}f(x,y)$$
\end{example}  
    
    Observe that theorem \ref{limmultirules} allows us to evaluate the limits, except when we possibly divide by zero.
    
  \begin{motivating}
  When do limits of multivariable functions not exist?
  \end{motivating}  
    
    \begin{definition}
    Let $X \subset \R^n$.  We say that a point $\bm{p} \in \R^n$ is a \textbf{limit point of} $X$ if there is a sequence $\{\bm{a_n}\}$ contained inside $X$ such that $\{\bm{a_n}\}$ converges to $\bm{p}$.    
    \end{definition}
        
    
    \begin{theorem}
     Let $X \subset \R^n$, $f : X \to \R^m$ a function, and $\bm{a}$ a limit point of $X$.  Then the following statements are equivalent:
     
     \begin{enumerate}
         \item $\lim_{\bm{x} \to \bm{a}}f(x) = \bm{b}$
         \item For every sequence $\{\bm{a_n}\}$ converging to $\bm{a}$ (with $\bm{a_n} \neq \bm{a}$), the sequence $\{f(\bm{a_n})\}$ converges to $\bm{b}$.
     \end{enumerate}
     
    \end{theorem}
    
    In other words, in order for a limit of a multivariable function to exist, it must yield the same value \textbf{along all possible approaches}.
    
    \begin{example}
    Show that $\lim_{(x,y) \to (0,0)} \frac{x^2}{x^2 + y^2}$ does not exist.
    
    Consider the paths $\bm{r_1}(t) = \langle 0, t \rangle$, and $\bm{r_2}(t) = \langle t, t \rangle$.
    \end{example}
    
    
    
    
    
    
    \begin{theorem}[Squeeze Theorem]
    Let $f(\bm{x})$, $g(\bm{x})$, and $h(\bm{x})$ be functions of $n$ variables such that $$\lim_{\bm{x} \to \bm{P}}f(\bm{x}) = L = \lim_{\bm{x} \to \bm{P}}h(\bm{x})$$
    
    If there exists $\delta > 0$ such that for all $\bm{x} \in B_\delta(\bm{P})$, we have that $$f(\bm{x}) \leq g(\bm{x}) \leq h(\bm{x})$$ 
    Then $\lim_{\bm{x} \to \bm{P}}g(\bm{x}) = L$.
    \end{theorem}
    
    \begin{example}
    Determine if the following limit exists: $$\lim_{(x,y) \to (0,0)} x\sin(\frac{1}{x^2 + y^2})$$ 
    \end{example}
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    

\subsection{Exercises}

\begin{problem}{limitsdotprod}

Let $\{\bm{a_n}\}$ and $\{\bm{b_n}\}$ be sequences in $\R^n$ that converge to $\bm{a}$ and $\bm{b}$, respectively.  Show that
      \begin{enumerate}
          \item $\lim_{n \to \infty}( \bm{a_n} \cdot \bm{b_n} ) = \bm{a} \cdot \bm{b}$
          \item $\lim_{n \to \infty}(||\bm{a_n}||) = || \bm{a}||$
      \end{enumerate}
\end{problem}

\begin{problem}{limitsmagnitude}

Let $\{\bm{a_n}\}$ be a sequence in $\R^n$ that converges to $\bm{a}$.  Show that
      \begin{enumerate}
          \item $\lim_{n \to \infty}(||\bm{a_n}||) = || \bm{a}||$
      \end{enumerate}

\end{problem}

\begin{problem}{prodquotient}

Why don't we have product and quotient rules for limits of two sequences $\{\bm{a_n}\}$, $\{\bm{b_n}\}$ in $\R^n$?

\end{problem}

\begin{problem}{scalarprodquotient}
Conjecture and prove product and quotient rules for a sequence $\{\bm{a_n}\}$ in $\R^n$ and a sequence $\{b_n\}$ in $\R$.
\end{problem}

\begin{problem}{lim0}
Show that $\lim_{(x,y) \to (0,0)} x\sin(\frac{1}{x^2 + y^2}) =0$.
\end{problem}

\begin{problem}{lim15}
Show that $\lim_{(x,y) \to (0,0)} \frac{y^2}{x^2 + y^2}$ does not exist.
\end{problem}

\begin{problem}{limsqueeze}
Use the squeeze theorem to compute $\lim_{(x,y) \to (0,0)} \frac{5xy^2}{x^2+y^2}$
\end{problem}

\begin{problem}{lim5}
Show that the following limit does not exist.
     $$\lim_{(x,y) \to (0,0)} \frac{xy^2}{x^2 + 3y^4}$$
\end{problem}

\begin{problem}{lim1}
$\lim_{(x,y) \to (1,2)} x^2+y$ exist? Prove or disprove your claim.
\end{problem}

\begin{problem}{lim2}
Does $\lim_{(x,y) \to (0,0)} \frac{xy}{x^2 + y^2}$ exist? Prove or disprove your claim.
\end{problem}

\begin{problem}{lim3}
Does $\lim_{(x,y,z) \to (0,0,0)} \frac{x + y+ z}{x^2 + y^2 +z^2}$ exist? Prove or disprove your claim.
\end{problem}

\begin{problem}{lim4}
Does $\lim_{(x,y,z) \to (0,0,0)} \frac{xyz}{x^3 + y^3 +z^3}$ exist? Prove or disprove your claim.
\end{problem}

\begin{problem}{sequenceofpoints}
Consider the function $f(x,y) = \left\{
		\begin{array}{ll}
			\cos(x) & \text{ if } x^2 + y^2 = \frac{1}{n} \text{ for some } n = 1, 2, 3, \dots \\
			0 & \text{ otherwise } 
		\end{array}
		\right.$
		
		Compute $\lim_{(x,y) \to (0,0)} f(x,y)$ or show that it does not exist.	
\end{problem}






\section{Continuity}\label{sec:continuity}

\subsection{Vector-valued functions}

\subsection{Multivariable functions}

$f : \R^m\to \R$

\subsection{General Multivariable functions}

$f : \R^m \to \R^n$


\subsection{Exercises}

\begin{problem}{cont0}
Consider the function  $f(x,y) = \left\{
		\begin{array}{ll}
			y^2\sin(\frac{1}{(x-4)^2 + y^2}) & \text{ if } (x,y) \neq (4,0) \\
			c & \text{ otherwise } 
		\end{array}
		\right.$
		
		Either show that there is a $c$ that will make $f$ continuous at $(4,0)$, or show that $f$ will not be continuous for any choice of $c$.
\end{problem}

\begin{problem}{cont1}
Consider the function $f(x,y) = \left\{
		\begin{array}{ll}
			\frac{x^2 + \sin(y)}{2x^2 + y +e^{xy}-1} & \text{ if } (x,y) \neq (0,0) \\
			1 & \text{ otherwise } 
		\end{array}
		\right.$
		
		Is $f$ continuous at $(0,0)$?
\end{problem}



\begin{problem}{cont2}
Consider the function  $f(x,y) = \left\{
		\begin{array}{ll}
			\frac{x^2  +\sin(xy)+y^2}{2x^2 + y^2} & \text{ if } (x,y) \neq (0,0) \\
			c & \text{ otherwise } 
		\end{array}
		\right.$
		
		Either show that there is a $c$ that will make $f$ continuous, or show that $f$ will not be continuous for any choice of $c$.
\end{problem}

\begin{problem}{cont3}
Consider the function $f(x,y) = \left\{
		\begin{array}{ll}
			\frac{y^5}{x^4 + y^4} & \text{ if } (x,y) \neq (0,0) \\
			1 & \text{ otherwise } 
		\end{array}
		\right.$
		
		Is $f$ continuous at $(0,0)$?
\end{problem}

\begin{problem}{cont4}
Consider the function  $g(x,y) = \left\{
		\begin{array}{ll}
			x^2\cos(\frac{1}{x^2 + y^2}) & \text{ if } (x,y) \neq (0,0) \\
			c & \text{ otherwise } 
		\end{array}
		\right.$
		
		Either show that there is a $c$ that will make $g$ continuous, or show that $g$ will not be continuous for any choice of $c$.
\end{problem}

\begin{problem}{cont5}
Consider the function  $f(x,y) = \left\{
		\begin{array}{ll}
			\frac{x^2}{x^2 + |y|^3} & \text{ if } (x,y) \neq (0,0) \\
			c & \text{ otherwise } 
		\end{array}
		\right.$
		
		Either show that there is a $c$ that will make $f$ continuous at $(0,0)$, or show that $f$ will not be continuous for any choice of $c$.
\end{problem}

\begin{problem}{cont6}

Consider the function  $f(x,y) = \left\{
		\begin{array}{ll}
			x\sin(\frac{1}{x^2 + y^2}) & \text{ if } (x,y) \neq (0,0) \\
			c & \text{ otherwise } 
		\end{array}
		\right.$
		
		Either show that there is a $c$ that will make $f$ continuous at $(0,0)$, or show that $f$ will not be continuous for any choice of $c$.

\end{problem}

\begin{problem}{cont7}
Consider the function  $g(x,y) = \left\{
		\begin{array}{ll}
			\frac{x^4  +2x^2y^2 +y^4}{x^2 + y^2} & \text{ if } (x,y) \neq (0,0) \\
			c & \text{ otherwise } 
		\end{array}
		\right.$
		
		Either show that there is a $c$ that will make $g$ continuous at $(0,0)$, or show that $g$ will not be continuous for any choice of $c$.
\end{problem}